New results in $\sfrac{k}{n}$ Power-Hours

### 2013-12-28 powerhour/paper/paper.tex
Getting close on paper

\usepackage{code}
\usepackage{cite}
% \usepackage{eclbkbox}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}
% \usepackage{fancyhdr}
% \usepackage{color}
% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
% \usepackage{everypage}

@@

% \usepackage{inlinebib}

@@

In order to efficiently enumerate BM5L, I improved the algorithm
again. Observe that the ``drink'' action associated with a rule
usually does not affect anything but the final outcome. The only
exception is that in the rule for \fullcup, the player must drink if
passing in a state other than \fullcup. Putting this aside for a
moment, note that we can just count the number of times each rule was
executed for each player, producing an $s$-dimensional vector $[ d_1,
  \ldots d_s ]$ for each player. That player is able to achieve many
different drink totals, specifically, $d_1 \times r_1 + \ldots + d_s
\times r_s$ where $r_i$ is 1 if the player should drink on that rule
and 0 if not. Simulating a game this way is even more like abstract
interpretation (we leave a concrete value free and compute a formula
rather than an integer), and allows us to evaluate many concrete games
at once. At the end, we simply plug in every legal value for $r_i$ for
each player and insert those games into the database. This last step
is where we must tend to the exception around \fullcup. We may not set
$r_0$ to 0 if the player ever passes \fullcup\ in a non-full state. A
very close approximation would be to insist that $r_0 = 1$ if the rule
in the \fullcup\ position does not output as \fullcup, but this is
inexact, as that rule may never be executed\footnote{We may be able to
  argue that in that case, there always exists another game that does
  not violate this condition. But I think it is simpler to just
  implement the rules.} Instead, during simulation we keep track of
whether each player ever actually passed a non-full cup from the
\fullcup\ state. If so, then we force $r_0 = 1$, which attends to this
special case.

@@

\section{Samping games}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{powerday3.pdf}
\end{center}\vspace{-0.1in}
\caption{All possible outcomes for the first two players in 3-player
  power days. These are the same games as the 3-player power hours,
  but at this scale makes it clear the groupings and their sparsity
  in the limit. Lines plotted from (0,0) to (60,*) and (*,60) show
  significant structure, but don't explain some of the interior points.
  (TODO: what are these games? what did the other player drink?)
  (TODO: Incorporate this into the text, perhaps after discussing
  the linear combination trick.)
}
\label{fig:powerday3}
\end{figure}

>>>>

\usepackage{url}
\usepackage{code}
% \usepackage{cite}
% \usepackage{eclbkbox}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}
% \usepackage{fancyhdr}
% \usepackage{color}
% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
% \usepackage{everypage}

@@

\usepackage{natbib}
% \usepackage{inlinebib}

@@

In order to efficiently enumerate BM5L, I improved the algorithm
again. Observe that the ``drink'' action associated with a rule
usually does not affect anything but the final outcome. The only
exception is that in the rule for \fullcup, the player must drink if
passing in a state other than \fullcup. Putting this aside for a
moment, note that we can just count the number of times each rule was
executed for each player, producing an $s$-dimensional vector $[ d_1,
  \ldots d_s ]$ for each player. That player is able to achieve many
different drink totals, specifically, $d_1 \times r_1 + \ldots + d_s
\times r_s$ where $r_i$ is 1 if the player should drink on that rule
and 0 if not. Simulating a game this way is even more like abstract
interpretation (we leave a concrete value free and compute a formula
rather than an integer), and allows us to evaluate many concrete games
at once. At the end, we simply plug in every legal value for $r_i$ for
each player and insert those games into the database. This last step
is where we must tend to the exception around \fullcup. We may not set
$r_0$ to 0 if the player ever passes \fullcup\ in a non-full state. A
very close approximation would be to insist that $r_0 = 1$ if the rule
in the \fullcup\ position does not output as \fullcup, but this is
inexact, as that rule may never be executed\footnote{We may be able to
  argue that in that case, there always exists another game that does
  not violate this condition. But I think it is simpler to just
  implement the rules.} Instead, during simulation we keep track of
whether each player ever actually passed a non-full cup from the
\fullcup\ state. If so, then we force $r_0 = 1$, which attends to this
special case. \label{sec:linear}

@@

\section{Sampling games}

To establish a result like ``BM5L cannot achieve $\langle 47, 27
\rangle$'' we really need to enumerate all the BM5L games. (Or make
some ad hoc proof of the fact, which seems quite difficult.) However,
to prove an existence result like ``BM7L can achieve $\langle 33, 49
\rangle$'' we only need to have a single example configuration that
produces that result. Therefore, we may be able to improve our bounds
on what is possible (or generate conjectures) by sampling random
configurations.

Sampling is actually much easier than enumeration. There is no need to
leave rules abstract. It is also easy to stop and restart because
there is no state other than the matrix of what we've found. I use the
SML {\tt textformat} library~\cite{textformat} to serialize and
deserialize the matrix (which then makes it easy to generate these
graphics in a separate program). There are a handful of interesting
aspects:

\paragraph{Generating a random configuration.} To generate a random
game, we can just fill in all of the slots (destination and cup state
for each rule, starting cup state) uniformly at random. Many of these
will be illegal, but they fail very quickly at runtime; a lazy and
pragmatic way to ``filter'' to legal game. It is not simply a matter
of generating all the permutations on $p \times s$ nodes, by the way.
Multiple cups can pass through the same player on cycles of different
periods, as long as they do not collide within the 60 steps, and
acyclic preludes (Figure~\ref{fig:solo31}) are important and useful.
For a uniformly random 2-player BM7L configuration, 29.23\% (measured
empirically) are legal. However, we will see later that we do not want
to spend so much time exploring configurations where one or both
players start without a cup; these are very limited. Therefore, the
configuration generator is biased towards producing a cup in the
starting states most of the time.

\paragraph{Symmetry.} We can get more bang for the buck by considering
some obvious symmetries. When a simulation finishes and we have an
outcome $\langle k_1, \ldots, k_p \rangle$, it is clear that any
permutation of $k_1 \ldots k_p$ is also achievable. We insert every
permutation of the drink counts into the database, along with the
permuted example configuration. Better still would be to only store
the outcomes in some normalized form (e.g. require that $k_1 \leq
\ldots \leq k_p$).

We already have exact results for two players in BM5L, so the next
uncharted territory is BM6L. The result after apparent convergence
appears in Figure~\ref{fig:samplebm6l}. The sampling procedure runs
for many hours before plateauing overnight with 95.54\% of the matrix
filled. This suggests that BM6L is not universal for two players, or
else the configurations for the missing cells are extremely rare.
% XXX never give up! It just jumped to 95.65.

This approach scales much better than enumeration and is efficient for
all sorts of generalizations (it works best when the dimensionality is
low---i.e., two players---and the expressiveness is high---i.e., many
cup states). Since we already know BM8L is universal, the remaining
open problem is BM7L, whose results are in
Figure~\ref{fig:samplebm7l}. This is tantalizingly close to a complete
matrix, with only 4 cells missing! It may be possible to solve these
two cases manually, though a proof of non-existence would be
difficult.

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min2players6states.png}
\end{center}\vspace{-0.1in}
\caption{ 25.6 billion samples of legal two-player BM6L
  configurations. Darker cells represent outcomes that occur more
  often; cells that are pure white never occurred and are likely to be
  unattainable. Note that the intensity represents the rank of
  occurrence, not the magnitude; in actuality, outcomes like $\langle
  0, 0 \rangle$ occur {\em much} more often than others.
  %
  95.54\% of the cells are filled.
  % XXX new version with more filled...!
}
\label{fig:samplebm6l}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min2players7states.png}
\end{center}\vspace{-0.1in}
\caption{ 22.1 billion samples of legal two-player BM7L
  configurations. Fewer configurations were sampled than in
  Figure~\ref{fig:samplebm6l} because they take somewhat longer than
  6-state games to simulate, and a smaller proportion of random games
  are legal.
  %
  All but four of the $(60 + 1)^2 = 3721$ cells are attainable: 
  Permutations of $\langle 11, 53 \rangle$ and $\langle 49, 53 \rangle$
  remain elusive. Note that 53 drinks was also unattainable in a solo
  BM7L power-hour; this may in some sense be the ``hardest'' number of
  shots to drink in \bmsl.
}
\label{fig:samplebm7l}
\end{figure}

@@

Some of the patterns are easy to explain. The top-left half of the
matrix is more populous than the bottom right, for example. This is
because we can bound the total number of drinks by $60 / c$, where $c$
is the number of cups active in the game (same as the number of cups
in the starting states). The top-left half is the region where this
sum is less than or equal to 30; in two player games, both players
must start with a cup in order to get an outcome in the bottom-right
half. We also see distinct clumps around 0, 15, 30, 45, 60; divisors
of 60. This is intuitive because the expressive power of BMML comes
from the ability to form cycles of cup states and drink on some
fraction of them. Clumps are formed around these values because of the
possibility of preludes leading into the cycles
(Figure~\ref{fig:solo31}) that either drink (adding to the total) or
don't (subtracting from it). Minor clumps form as echoes between the
major ones, because a player may participate in two cycles of
different length (Figure~\ref{fig:dual45}.

Of course, discretization effects compound and so the exact values of
cells are not neatly predictable. Moreover, clumps interfere by
overlapping; there are many different strategies for achieving
$\langle 33, 33 \rangle$. One way to make the basic structure more
visible is to extend the number of minutes that the game is played
for. Figure~\ref{fig:powerday3} shows the utterly unhealthy
three-player Power Day (BM3L). In it, the clumps become tiny dots,
but some relationship among them along lines is clear. Interior points
can probably be found as linear combinations of two of these lines;
we exploit that exact structure in Section~\ref{sec:linear}, in fact.

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{powerday3.pdf}
\end{center}\vspace{-0.1in}
\caption{All possible outcomes for the first two players in 3-player
  power days. These are the same games as the 3-player power hours,
  but at this scale makes it clear the groupings and their sparsity in
  the limit. Lines plotted from $\langle 0,0 \rangle$ to $langle 60,
  k_2 \rangle$ and $\langle k_1, 60 \rangle$ show significant
  structure, but don't explain some of the interior points. These are
  probably games where a player participates in two cycles of
  different periods.
}
\label{fig:powerday3}
\end{figure}
